---
title: "p8157_hw1_jsg2145"
author: "Jared Garfinkel"
date: "9/25/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(Hotelling)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	message = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  scipen = 999
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

# Part A

## Problem 1

B is a 2 x 6 matrix of group means. 

### Part a

```{r}
A = matrix(c(1, -1), ncol = 2, byrow = TRUE)
A

C = diag(6)
C
```

### Part b

```{r}
A = matrix(c(1, -1), ncol = 2, byrow = TRUE)
A

C = matrix(c(-1, 0, 0, 0, 0,
             1, -1, 0, 0, 0,
             0, 1, -1, 0, 0,
             0, 0, 1, -1, 0,
             0, 0, 0, 1, -1,
             0, 0, 0, 0, 1), ncol = 5, byrow = TRUE)
C
```

### Part c

```{r}
A = matrix(c(-1, 1), ncol = 2, byrow = TRUE)
A

C = matrix(c(0,0,0,0,1,-1))
C
```

# Part B

## Problem 1

Exploratory Data Analysis of blood lead levels.

```{r}
dat = read.table("./data/TLC.dat") %>% 
  janitor::clean_names() %>% 
  rename("id" = v1,
         "group" = v2,
         "week0" = v3,
         "week1" = v4,
         "week4" = v5,
         "week6" = v6) %>% 
  mutate(group = recode(group,
                        "P" = 0,
                        "A" = 1))
```

```{r, results = "hide"}
dat %>% 
  pivot_longer(cols = week0:week6, names_to = "t", values_to = "mg_dL") %>% 
  ggplot(aes(x = t, y = mg_dL, group = id)) +
  geom_path(aes(color = id)) +
  theme(legend.position = "none")
```

It appears that those in one group have a response to treatment after the first week, but that the effects begin to wear off by week 6.

```{r}
dat %>% 
  pivot_longer(cols = week0:week6, names_to = "t", values_to = "mg_dL") %>% 
  ggplot(aes(x = t, y = mg_dL, group = id)) +
  geom_path(aes(color = group)) +
  theme(legend.position = "none")
```



```{r, results = "hide", echo = FALSE}
test1 = hotelling.test(.~group, data = dat[,-1])
test1
```

```{r, results = "hide", echo = FALSE}
cmat = matrix(c(-1, 1, 0, 0, 0, -1, 1, 0, 0, 0, -1, 1), ncol = 4, byrow = TRUE)
cmat

data2 = t(dat[,-c(1:2)]) 
data2

data3 = cmat %*% data2
data3 = t(data3)
data3 = cbind(dat[,2], data3) %>% 
  data.frame()
data4 = data3 %>% 
  mutate(across(.cols = c(X2:X4), .fns = as.numeric))
```

```{r, results = "hide", echo = FALSE}
test.contrast = hotelling.test(.~X1, data = data4)
test.contrast
```

# Problem 2

```{r}
df = read.table("./data/ZERBE2.DAT") %>% 
  janitor::clean_names() %>% 
  rename("group" = v1,
         "id" = v2,
         "t0" = v3,
         "t0.5" = v4,
         "t1" = v5,
         "t1.5" = v6,
         "t2" = v7,
         "t3" = v8)
```

## Part a

Test the group means.

```{r}
test.ph = hotelling.test(.~group, data = df[,-2])
test.ph
```

Since the p-value is `r test.ph$pval` we can reject the null. There is evidence of a difference of means among groups.

## Part b

Test the parellelism.

```{r,results="hide"}
cmat = matrix(c(-1, 1, 0, 0, 0, 0, 
                0, -1, 1, 0, 0, 0, 
                0, 0, -1, 1, 0, 0, 
                0, 0, 0, -1, 1, 0, 
                0, 0, 0, 0, -1, 1), ncol = 6, byrow = TRUE)
cmat

df2 = t(df[,-c(1:2)]) 
df2
# 6 x 33

df3 = cmat %*% df2
df3 = t(df3)
df3 = cbind(df[,1], df3) %>% 
  data.frame()
```

```{r}
test.contrast2 = hotelling.test(.~X1, data = df3)
test.contrast2
```

Since the p-value is `r test.contrast2$pval` we may reject the null indicating that there is a difference in the slopes between time points among groups.

```{r, results = "hide", echo = FALSE}
cmat3 = matrix(c(1, 0, 0, 0, 0, 
                 0, 1, 0, 0, 0, 
                 0, 0, 1, 0, 0,
                 0, 0, 0, 1, 0,
                 0, 0, 0, 0, 1,
                 -1, -1, -1, -1, -1), ncol = 5, byrow = TRUE)



# unit = matrix(c(1, 1))

df_dat = t(df[,-c(1:2)])
cmat3 = t(cmat3)

df_dat2 = cmat3 %*% df_dat
df_dat3 = t(df_dat2)
df_dat3 = cbind(df[,1], df_dat3) %>% 
  data.frame()

test.contrast3 = hotelling.test(.~X1, data = df_dat3)
test.contrast3
```

## Part c

Test the difference in group means between the last two time points.

```{r}
df_final = df %>% 
  select(t2, t3) %>% 
  t()

cmat_final = matrix(c(1, -1), ncol = 2, byrow = TRUE)

df_final2 = cmat_final %*% df_final %>% 
  t()

df_final3 = cbind(df[,1], df_final2) %>% 
  data.frame()

test.contrast3 = hotelling.test(.~X1, data = df_final3)
test.contrast3
```

The pvalue is `r test.contrast3$pval`, indicating we do not reject the null, and there is not evidence of a difference in the time points between groups for time t = 2 and time t = 3..

## Exploratory Data Analysis

Exploratory data analysis does not appear to show much of a difference between groups at first.

```{r}
df %>%
  mutate(group = factor(group, labels = c("control" = 1, "obese" = 2))) %>% 
  pivot_longer(cols = t0:t3, names_to = "t", values_to = "plasma") %>% 
  ggplot(aes(x = t, y = plasma, group = id)) +
  geom_path(aes(color = group)) +
  scale_color_discrete(labels = c("control", "obese"))
```

In a side-by-side plot, it appears that perhaps both groups decrease in plasma phosphate levels over time. The control group may have more a "U" shape, where the phosphate levels decrease and then increase. On the other hand, it appears to take longer for those in the obese group to reach a valley in the phosphate levels. 

As with the test for parallelism, one may test the slopes between individuals timepoints. Furthermore, it may be of interest to determine at which time point there is a difference in group means.

```{r}
 df %>%
  mutate(group = factor(group, labels = c("control" = 1, "obese" = 2))) %>% 
  pivot_longer(cols = t0:t3, names_to = "t", values_to = "plasma") %>% 
  ggplot(aes(x = t, y = plasma, group = id)) +
  geom_path(aes(color = group)) +
  scale_color_discrete(labels = c("control", "obese")) +
  facet_grid(.~group)
```



